Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-06-18 paper.tex
Updates from Overleaf

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$. 
Similar to the decoder $\psi$, we use the vector representations $f_t \in \mathbb{R}^k$ computed by the LSTM encoder after processing $X_{-M\dots t}$ for $t = -M, \dots, -1$, and then compute predictions using the softmax rule
\begin{equation}
	P_\eta(X_t = s_i|X_{1\dots t-1}, Z) \propto \exp((W_{o'} f_t)_i)
\end{equation}
where $W_{o'} \in \mathbb{R}^{|S|\times k}$ is another parameter matrix.

Because we consider stationary processes, we have 
that the cross-entropy under $P_\eta$ of $\finitefuture$ is equal to the cross-entropy of $\finitepast$ under the same encoding distribution: $\mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture)\right] = - \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$. \rljf{I see that this is true if $P_\eta$ is the true marginal but what if it is approximate? Could $P_\eta$ end up non-stationary even when it comes from an encoder fit to a stationary distribution? Or is the stationarity of $P_\eta$ guaranteed because the encoder LSTM is autoregressive?} \mhahn{If $X_t$ is stationary, then the distributions of $\finitepast$ and $\finitefuture$ are the same, so these expectations are the same. Is there a way to make this more transparent?}
Using this observation, we estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~\cite{mcallester2018formal}:
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture) - \log P_\psi(\finitefuture | Z)\right],
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

@@

In the neural networks literature, the most commonly used method using variational bounds similar to Equation~(\ref{eq:bound}) is the Variational Autoencoder \citep{kingma-auto-encoding-2014, bowman-generating-2016}, which corresponds to the setting where $\lambda=1$ and the predicted output is equal to the observed input.
The $\beta$-VAE \citep{higgins2017beta}, a variant of the Variational Autoencoder, uses $\lambda > 1$ (whereas the Predictive Rate-Distortion objective~(\ref{eq:ib}) uses $\lambda \in (0,1)$), and has been linked to the Information Bottleneck by \cite{burgess2018understanding}.\footnote{As discussed in Section~\ref{sec:background}, our model becomes degenerate with $\lambda \ge 1$. Surprisingly, $\beta$-VAE operates in exactly this parameter regime, although }

@@

\paragraph{OCF} As discussed in Section~\ref{sec:OCF}, OCF is affected by overfitting, 
and will systematically overestimate the predictiveness achieved at a given rate~\citep{still-optimal-2010}.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and predictiveness on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|\finitepast)$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCF and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{prop:kl-bound}:
\begin{equation}\label{eq:rate-OCF}
\frac{1}{N} \sum_{\finitepast \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|\finitepast)}{s(Z)}
\end{equation}
where $P(Z|\finitepast)$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-OCF}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{prop:mi-bound}:
\begin{equation}
\frac{1}{N} \sum_{\finitepast, \finitefuture \in \text{Test Data}} \E_{Z \sim P(\cdot|\finitepast)} \log P(\finitefuture|Z).
\end{equation}
where $P(\finitefuture|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to Propositions~\ref{prop:mi-bound} and \ref{prop:kl-bound}, these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

The relevance of prediction to human language processing makes the \emph{difficulty} of prediction another interesting aspect of language complexity:
Predictive Rate-Distortion describes how much memory of the past humans need to maintain to predict future words accuractely.
Beyond the entropy rate, it thus forms another important aspect of linguistic complexity.

@@

Training used Early Stopping using the development set, as described in Section~\ref{sec:parameter-estimation}.
Models were trained on a TITAN Xp graphics card.

\mhahn{TODO number of epochs, training time comparable to OCF}
On Even Process and Random Insertion Process, OCF takes at most a few seconds at $M \leq 5$.
On English word-level modeling, training took a median number of 9 epochs (max 467 epochs) and 5 minutes (max 126 minutes).

@@

\begin{figure*}
\centering
\begin{tabular}{c}
\textbf{Past:} within minutes after the stock market closed friday , i called sen. bill bradley of new \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_0.png}\\
\textbf{Past:} asset you own would be worth more if the government suddenly announced that if you sold \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_20.png}\\
\textbf{Past:} he thinks , when the white house is forced to ask for higher taxes to meet \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_31.png} \\
\textbf{Past:} to be dealt with by the oct. 15 deadline , and these senate democrats refused to \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_6.png} \\
\textbf{\textcolor{red}{- - - -}}\ \ Low rate ($\log \frac{1}{\lambda}=1.0$)\ \ \ \ \ \ 
\textbf{\textcolor{olive}{-- - --}}\ \ Median rate ($\log \frac{1}{\lambda}=3.0$)\ \ \ \ \ \ 
\textbf{\textcolor{blue}{------}}\ \ High rate ($\log \frac{1}{\lambda}=5.0$)
\end{tabular}

	\caption{Four example outputs from English word-level modeling, with low rate ($\log \frac{1}{\lambda} = 1$; red, dotted), medium rate ($\log \frac{1}{\lambda} = 3$; green, dashed), high rate ($\log \frac{1}{\lambda} = 5$; blue, solid). For each sample, we provide the prior context $\finitepast$ (top), and the per-word cross-entropies (in nats) on the future words $\finitefuture$ (bottom). %\rljf{Figure needs y axis label and a legend to explain the meanings of the different colors}
	}\label{fig:samples-ptb}
\end{figure*}

>>>>

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$. $P_\eta$ has support over sequences of length $M$.
Similar to the decoder $\psi$, we use the vector representations $f_t \in \mathbb{R}^k$ computed by the LSTM encoder after processing $X_{-M\dots t}$ for $t = -M, \dots, -1$, and then compute predictions using the softmax rule
\begin{equation}
	P_\eta(X_t = s_i|X_{1\dots t-1}, Z) \propto \exp((W_{o'} f_t)_i),
\end{equation}
where $W_{o'} \in \mathbb{R}^{|S|\times k}$ is another parameter matrix.

Because we consider stationary processes, we have 
that the cross-entropy under $P_\eta$ of $\finitefuture$ is equal to the cross-entropy of $\finitepast$ under the same encoding distribution: $\mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture)\right] = - \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$. 
Using this observation, we estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~\cite{mcallester2018formal}:
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture) - \log P_\psi(\finitefuture | Z)\right],
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

@@

In the neural networks literature, the most commonly used method using variational bounds similar to Equation~(\ref{eq:bound}) is the Variational Autoencoder \citep{kingma-auto-encoding-2014, bowman-generating-2016}, which corresponds to the setting where $\lambda=1$ and the predicted output is equal to the observed input.
The $\beta$-VAE \citep{higgins2017beta}, a variant of the Variational Autoencoder, uses $\lambda > 1$ (whereas the Predictive Rate-Distortion objective~(\ref{eq:ib}) uses $\lambda \in (0,1)$), and has been linked to the Information Bottleneck by \cite{burgess2018understanding}.

@@

\paragraph{OCF} As discussed in Section~\ref{sec:OCF}, OCF is affected by overfitting, 
and will systematically overestimate the predictiveness achieved at a given rate~\citep{still-optimal-2010}.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and predictiveness on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|\finitepast)$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCF and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{prop:kl-bound}:
\begin{equation}\label{eq:rate-OCF}
\frac{1}{N} \sum_{\finitepast \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|\finitepast)}{s(Z)},
\end{equation}
where $P(Z|\finitepast)$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-OCF}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{prop:mi-bound}:
\begin{equation}
\frac{1}{N} \sum_{\finitepast, \finitefuture \in \text{Test Data}} \E_{Z \sim P(\cdot|\finitepast)} \log P(\finitefuture|Z).
\end{equation}
where $P(\finitefuture|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to Propositions~\ref{prop:mi-bound} and \ref{prop:kl-bound}, these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

The relevance of prediction to human language processing makes the \emph{difficulty} of prediction another interesting aspect of language complexity:
Predictive Rate-Distortion describes how much memory of the past humans need to maintain to predict future words accurately.
Beyond the entropy rate, it thus forms another important aspect of linguistic complexity.

@@

Training used Early Stopping using the development set, as described in Section~\ref{sec:parameter-estimation}.
Models were trained on a TITAN Xp graphics card.
On Even Process and Random Insertion Process, NPRD took a median of 10 minutes (max 48 minutes).
OCF took less than one minute at $M \leq 5$; however, it does not scale to larger values of $M$.
On English word-level modeling, training took a median number of 9 epochs (max 467 epochs) and 5 minutes (max 126 minutes).

@@

\definecolor{darkgreen}{rgb}{0.33, 0.6, 0.18}

\begin{figure*}
\centering
\begin{tabular}{c}
\textbf{Past:} within minutes after the stock market closed friday , i called sen. bill bradley of new \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_0.png}\\
\textbf{Past:} asset you own would be worth more if the government suddenly announced that if you sold \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_20.png}\\
\textbf{Past:} he thinks , when the white house is forced to ask for higher taxes to meet \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_31.png} \\
\textbf{Past:} to be dealt with by the oct. 15 deadline , and these senate democrats refused to \\
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_6.png} \\
\textbf{\textcolor{red}{- - - -}}\ \ Low rate ($\log \frac{1}{\lambda}=1.0$)\ \ \ \ \ \ 
\textbf{\textcolor{darkgreen}{-- - --}}\ \ Medium rate ($\log \frac{1}{\lambda}=3.0$)\ \ \ \ \ \ 
\textbf{\textcolor{blue}{------}}\ \ High rate ($\log \frac{1}{\lambda}=5.0$)
\end{tabular}

	\caption{Four example outputs from English word-level modeling, with low rate ($\log \frac{1}{\lambda} = 1$; red, dotted), medium rate ($\log \frac{1}{\lambda} = 3$; green, dashed), high rate ($\log \frac{1}{\lambda} = 5$; blue, solid). For each sample, we provide the prior context $\finitepast$ (top), and the per-word cross-entropies (in nats) on the future words $\finitefuture$ (bottom).
	}\label{fig:samples-ptb}
\end{figure*}
